% This file was converted from HTML to LaTeX with
% Tomasz Wegrzanowski's <maniek@beer.com> gnuhtml2latex program
% Version : 0.1
\documentclass{article}
\begin{document}



  
  
    
      
      
      
    
  


\section*{Chapter4}


[\textit{Previous Chapter}]
[\textit{Contents}]
[\textit{Next Chapter}]


\section*{Basic Pitch Translations}



Many musical processes entail some sort of data translation in which
one form of representation is transformed into another form of representation.
There are innumerable examples of such musical "translations."
For example, we might rewrite guitar tablatures as notated pitches.
Pitches might be translated to tonic-solfa syllables.
Scale-degrees might be analyzed as Roman-numeral harmonies.
Figured bass notations might be "realized" as pitches.
Successive pitches might be characterized as melodic intervals.
Intervals might be rewritten as semitone distances.
Pitch-class sets might be transformed to interval-class vectors.

\par 
In this chapter we introduce some basic musical translations.
We will limit this initial discussion to translating between
various representations related to pitch.
These will include French, German, and international pitch designations,
frequency, semitones, cents, solfège, scale-degree and MIDI pitch
representations.
In later chapters we will describe additional types of translations.


\subsection*{ISO Pitch Representation}

\par 
The best-known system for representing equally-tempered pitches
is the International Standards Organization (ISO) format consisting of
a letter (A-G) followed by an optional sharp or flat,
followed by an octave number.
According to the ISO representation middle C is designated C4.
Octave numbers change between B and C so that the B a semitone below C4 is B3.
Humdrum provides a predefined ISO-like representation called
\texttt{**pitch}
illustrated below.
Here we see an ascending chromatic scale in the left spine,
with a concurrent descending chromatic scale in the right spine:

\texttt{**pitch**pitch
C4C5
C\#4B4
D4Bb4
D\#4A4
E4Ab4
F4G4
F\#4Gb4
G4F4
G\#4E4
A4Eb4
A\#4D4
B4Db4
C5C4
*-*-}

Notice that only upper-case letters are used for pitch-name and
that the flat is represented by the lower-case letter `\texttt{b}'.
The small letter `\texttt{x}' can be used to indicate double-sharps
and the double-flat is represented by two small successive letters `\texttt{bb}'.
If a given pitch deviates from equal temperament,
cents deviation can be indicated by trailing integers preceded
by either a plus sign (tuned sharp) or a minus sign (tuned flat).
The unit of the
\textit{cent}
is one one-hundredth of a semitone,
so the distance between C4 and C\#4 is 100 cents.
For example, the pitch corresponding to a tuning of A-435 Hz is
19 cents flat compared with the standard A-440,
hence it is represented in \texttt{**pitch} as:

\par 

\texttt{A4-19}



\par 
Other pitch representations (such as \texttt{**kern}) can be
translated to the ISO-inspired \texttt{**pitch} representation by invoking the
\textbf{pitch}
command.
For example, consider the following \texttt{**kern} input:

\texttt{**kern
=1
4g
4g\#
4a
4cc
=2
*-}

It can be translated to this \texttt{**pitch} output:

\texttt{**pitch
=1
G4
G\#4
A4
C5
=2
*-}

using the following command:

\par 

\texttt{pitch filename}


\par 
Notice that the
\texttt{**pitch}
representation uses the same
system for representing barlines as \texttt{**kern}.
In fact, all of the pitch-related representations described
in this chapter make use of the so-called `common system' for
representing barlines.


\subsection*{German Tonhöhe}

\par 
The German system of pitch designations (\textit{Tonhöhe})
is available in the Humdrum
\texttt{**Tonh}
representation.
The German system is similar to the ISO system with the following
exceptions.
The pitch letter names include A,B,C,D,E,F,G,H and S.
Sharps and flats are indicated by suffixes: `\texttt{is}' for sharps
and `\texttt{es}' for flats,
hence `\texttt{Cis}' for C-sharp and `\texttt{Ges}' for G-flat.
Suffixes are repeated for double and triple sharps and flats.
Special exceptions include the following: `\texttt{B}' for B-flat,
`\texttt{H}' for B-natural, `\texttt{Heses}' for B double-flat
(rather than `\texttt{Bes}'), and `\texttt{As}' and `\texttt{Es}'
rather than `\texttt{Aes}' or `\texttt{Ees}'.
`\texttt{S}' is a short-hand for `\texttt{Es}' (E-flat).
As in the ISO pitch system, octaves are indicated by integer
numbers with middle C represented as C4.

\par 
Although modern German practice has gravitated toward the
ISO system, the traditional German system for
representing pitches remains important in historically
related studies, such as searching for `B-A-C-H' and
the pitch signature used by Dmitri Shostakovich (`D-S-C-H').

\par 
Data in the
\texttt{**pitch}
or
\texttt{**kern}
representations
can be translated to
\texttt{**Tonh}
via the
\textbf{tonh}
command:

\par 

\texttt{tonh filename}



\subsection*{French Solfège}

\par 
The common French system for pitch naming uses a so-called
"fixed-do" method of diatonic pitch designations:
\textit{do, ré, mi, fa, sol, la}
and
\textit{si}
(rather than \textit{ti}), where
\textit{do}
corresponds to the English/German `C'.
In the Humdrum
\texttt{**solfg}
representation, solfège
pitch names are used.
Flats (\textit{bémol}) and sharps (\textit{dièse}) are
abbreviated
\textit{b}
and
\textit{d}
respectively.
When accidentals are encoded, the tilde character (\texttt{~})
is encoded following the solfège syllable and before
the accidental.
Double and triple sharps and flats are encoded via repetition.
Hence, `do dièse' (\texttt{do~d}) for C-sharp, `la bémol'
(\texttt{la~b}) for A-flat, `sol double-dièse' (\texttt{sol~dd})
for G double-sharp, `si double-bémol' (\texttt{si~bb})
for B double-flat, and so on.
As with the German and ISO pitch representations, octave is designated
by integers with \texttt{do4} representing middle C.


\subsection*{Frequency}

\par 
For acoustic-related applications it may be helpful to
translate to frequency.
The Humdrum
\texttt{**freq}
representation can be used to
represent frequencies for either pure or complex tones.
Frequencies are encoded in
\textit{hertz}
(abbreviated \textit{Hz})
where 440 Hz means 440 cycles per second.
In the \texttt{**freq} representation frequencies may be
specified as integer or real values (with a decimal point).


\subsection*{Cents}

\par 
The
\texttt{**cents}
representation provides a means for
representing pitches in absolute units with respect to middle C (= 0 cents).
In the \texttt{**cents} representation,
all pitches are represented with respect to this
reference.
Thus C\#4 is represented by the number 100,
A4 is represented by 900, and A3 is represented by -300.
As in the case of \texttt{**freq}, cents may be specified as integer
numbers or as real values (with a decimal point).


\subsection*{Semitones}

\par 
A related pitch representation is
\texttt{**semits}.
In this case, all pitches are represented in numerical semitones
with respect to middle C (= 0 semits).
An ascending chromatic scale beginning on C4 would be represented
by the ascending integers from 0 to 12.
Pitches below middle C are represented by negative values.
Fractional values can be represented using decimal points.


\subsection*{MIDI}

\par 
Another way of representing pitch is provided by the Humdrum
\texttt{**MIDI}
representation.
This representation closely mimics
the commercial MIDI specification.
The \texttt{**MIDI} representation allows MIDI inputs and outputs
to be exported or imported by various Humdrum tools.
A complete description of \texttt{**MIDI} will be given in
Chapter 7.


\subsection*{Scale Degree -$\,$- \textit{**solfa} and \textit{**deg}}

\par 
Two different Humdrum representations are provided to
describe scale-degree related information:
\texttt{**deg}
and
\texttt{**solfa}.
Both of these representations emphasize slightly different
aspects of scale-degree information.
Both representations assume some established or pre-defined
tonal center or tonic pitch.

\par 
The \texttt{**solfa} representation represents pitch according
to tonic solfa syllables.
Pitches are designated by the syllables
\textit{do, re, mi, fa, so, la}
and
\textit{ti}
or their chromatic alterations as indicated in the following
table:

basicraisedlowered

do (\textit{doe})di (\textit{dee})de (\textit{day})
re (\textit{ray})ri (\textit{ree})ra (\textit{raw})
mi (\textit{me})my (\textit{my})me (\textit{may})
fa (\textit{fah})fi (\textit{fee})fe (\textit{fay})
so (\textit{so})si (\textit{see})se (\textit{say})
la (\textit{la})li (\textit{lee})le (\textit{lay})
ti (\textit{tee})ty (\textit{tie})te (\textit{tay})


\textit{Summary of \textbf{solfa} Signifiers}

The
\texttt{**deg}
representation identifies scale-degrees by
the numbers 1 (tonic) to 7 (leading-tone).
These values may be chromatically altered by raising (+)
or lowering (-).
The
\textit{amount}
of chromatic alteration is not indicated;
for example, both a raised supertonic and a doubly-raised supertonic
are represented as \texttt{2+}.
A lowered dominant is represented as \texttt{5-}.

\par 
The
\texttt{**solfa}
representation differs from
\texttt{**deg}
in that pitches are represented without regard to major or minor
\textit{mode.}
For example, in the key of C major, \texttt{**deg}
will characterize A-flat as a lowered sixth scale
degree (\texttt{6-}),
whereas the same pitch will be a normal (unaltered) sixth scale
degree in the key of C minor (\texttt{6}).
In the case of \texttt{**solfa}, the A-flat will be represented
as \texttt{le} -$\,$- whether or not the key is C major or C minor.
Like \texttt{**deg}, the amount of chromatic alteration is not
represented in \texttt{**solfa}.
Once a pitch is raised, raising it further will not
change the representation.
For example, if the tonic is B-flat, then both B-natural and B-sharp
will be represented by \texttt{di} in the \texttt{**solfa} representation.

\par 
In the case of the minor mode, \texttt{**deg} characterizes scale
degrees with respect to the
\textit{harmonic minor}
scale only.

\par 
Another difference between
\texttt{**solfa}
and
\texttt{**deg}
is that the \texttt{**deg} representation provides a way for encoding
\textit{melodic approach.}
The caret (\texttt{\^{}}) denotes an ascending melodic approach to
the current note, whereas the lower-case letter \texttt{v} denotes
a descending melodic approach.
Repeated pitches carry no melodic approach signifier.

\par 
Some of the differences between the \texttt{**solfa} and
\texttt{**deg} representations are illustrated in Example 4.1.
(The corresponding
\texttt{**kern}
representation is given in the first spine.)
Notice that \texttt{**solfa} does not encode any octave information.
The \texttt{**deg} representation does not encode the octave of
the starting pitch, but it does indicate contour information
using the caret (\texttt{\^{}}) for ascending and the lower-case
\texttt{v} for descending pitches.
Notice also the different ways of characterizing accidentals.
\\
\\
\textbf{Example 4.1}


\texttt{!! Comparisonof pitch-relatedrepresentations.
\texttt{**kern**solfa**deg
\texttt{*M2/4*M2/4*M2/4
\texttt{*c:*c:*c:
\texttt{8.ccdo1
\texttt{16ddre\^{}2
\texttt{=1=1=1
\texttt{8.ee-me\^{}3
\texttt{16ddrev2
\texttt{4eenmi\^{}3+
\texttt{=2=2=2
\texttt{8rrr
\texttt{8b-tev7-
\texttt{8anlav6+
\texttt{8ccdo\^{}1
\texttt{=3=3=3
\texttt{2bntiv7
\texttt{======
\texttt{*-*-*-}


\par 

\subsection*{Pitch Translations}


\par 
Humdrum provides a number of commands for translating
between the various pitch-related representations described above.
Typically, the command name is the same as the name of the output
representation.
For example, translating to
the
\texttt{**solfg}
representation can be accomplished with:

\par 

\texttt{solfg inputfile > outputfile}



\par 
Translating to the German
\texttt{**Tonh}
representation:

\par 

\texttt{tonh inputfile > outputfile}


\par 
Translating to ISO
\texttt{**pitch}:

\par 

\texttt{pitch inputfile > outputfile}




\par 
Similarly, the
\textbf{freq}
command translates pitch-related inputs to the
\texttt{**freq}
representation, the
\textbf{cents}
command translates appropriate inputs to the
\texttt{**cents}
representation, and so on.


\par 
In a few cases, the command names are slightly modified.
All Humdrum command names employ lower-case letters only,
so
\texttt{**MIDI}
output is generated by the
\textbf{midi}
command (rather than the \textbf{MIDI} command),
and \texttt{**Tonh} output is generated by the
\textbf{tonh}
command.

\par 
Examples 4.2 and 4.3 compare several parallel
representations of the same pitch-related information.
In both examples, the pitch information has been derived from the
\texttt{**kern} data shown in the left-most spine.
The duration information in the \texttt{**kern} data is not
available in the other representations.
However, the `common system' for barlines is used throughout.

\par 
Example 4.2 shows four pitch naming systems:
ISO pitch, German Tonhöhe, French solfège,
as well as \texttt{**kern}.
Notice the different ways of treating accidentals such as the
D-sharp and B-flat.
Also note the German use of H for B-natural.
\\
\\
\textbf{Example 4.2}


\texttt{!! Comparisonof pitch-relatedrepresentations.
\texttt{**kern**pitch**Tonh**solfg
\texttt{*M2/4*M2/4*M2/4*M2/4
\texttt{*C:*C:*C:*C:
\texttt{8.ccC5C5do5
\texttt{16ddD5D5re5
\texttt{=1=1=1=1
\texttt{8.eeE5E5mi5
\texttt{16dd\#D\#5Dis5re~d5
\texttt{4eeE5E5mi5
\texttt{=2=2=2=2
\texttt{8rrrr
\texttt{8b-Bb4B4si~b4
\texttt{8aA4A4la4
\texttt{8cC4C4do4
\texttt{=3=3=3=3
\texttt{2bnB4H4si4
\texttt{========
\texttt{*-*-*-*-}

In Example 4.3 four of the more technical representations are
illustrated, including frequency and cents.
Notice that the
\texttt{**MIDI}
representation uses key-numbers to represent pitch:
key-on events are indicated by positive integers (between two slashes)
and key-off events are indicated by negative integers.
More detail concerning \texttt{**MIDI} is given in
Chapter 7.
\\
\\
\textbf{Example 4.3}


\texttt{!! Comparisonof pitch-relatedrepresentations(continued).
\texttt{**kern**semits**cents**MIDI**freq
\texttt{*M2/4*M2/4*M2/4*Ch1*M2/4
\texttt{*C:*C:*C:*M2/4*C:
\texttt{****C:*
\texttt{8.cc121200/72/523.25
\texttt{16dd141400/-72/ /74/587.33
\texttt{=1=1=1=1=1
\texttt{8.ee161600/-74/ /76/659.26
\texttt{16dd\#151500/-76/ /75/622.25
\texttt{4ee161600/-75/ /76/659.26
\texttt{=2=2=2=2=2
\texttt{8rrr/-76/r
\texttt{8b-101000/70/466.16
\texttt{8a9900/-70/ /69/440.00
\texttt{8c00/-69/ /60/261.63
\texttt{=3=3=3=3=3
\texttt{2bn111100/-60/ /71/493.88
\texttt{==========
\texttt{.../-71/.
\texttt{*-*-*-*-*-}

Not all of the above pitch-related representations can be translated
directly from one to another.
Table 4.1 shows the possible translations supported by Humdrum Release 2.0
commands.
The input representations are listed from right to left.
Under each column, those commands that will translate
\textit{from}
the given format are identified.
For example, the
\texttt{**cents}
representation can be translated
to \texttt{**freq},
\texttt{**kern},
\texttt{**pitch},
\texttt{**semits},
\texttt{**solfg},
and \texttt{**tonh}.
Notice that \texttt{**deg} data cannot be translated to any other
format since \texttt{**deg} representations do not encode absolute
pitch height.
Note also that when translating to the \texttt{**kern} representation,
only pitch-related information is translated:
duration, articulation marks, and other \texttt{**kern} signifiers are
not magically generated.
\\
\\
\textbf{Table 4.1}
\\
\\

\textbf{Input Representation}



**cents**deg**freq**kern**MIDI**pitch**semits**solfa**solfg**Tonh

cents¤¤¤¤¤¤
cocho¤
deg¤¤¤¤
freq¤¤¤¤¤¤¤
kern¤¤¤¤¤¤¤¤
midi¤
pitch¤¤¤¤¤¤¤¤
semits¤¤¤¤¤¤¤
solfa¤¤¤¤
solfg¤¤¤¤¤¤¤
tonh¤¤¤¤¤¤¤¤



\subsection*{Transposition Using the \textit{trans} Command}


\par 
A common pitch-related manipulation is transposition.
The 
\textbf{trans}
command has the user specify a
\textit{diatonic offset}
and a \textit{chromatic offset}.
The diatonic offset affects the pitch-letter name used to spell
a note.
The chromatic offset affects the number of semitones shifted
from the original pitch height.
The two types of offset are completely independent of each other.
For common transpositions, both the diatonic and chromatic offsets
will need to be specified.
For example, in transposing up a minor third (e.g. C to E-flat),
the diatonic offset is `up two pitch-letter names,'
and the chromatic offset is `up three semitones.'
The appropriate command invocation is:

\par 

\texttt{trans -d +2 -c +3 input > output}


\par 
The diatonic offset can be a little confusing because
traditional terminology labels perfect unisons by the number 1 (e.g. P1)
rather than zero.
So transposing up a perfect fifth involves a diatonic offset of +4
letter names, and a chromatic offset of +7 semitones:

\par 

\texttt{trans -d +4 -c +7 input > output}



\par 
We can transpose without changing the diatonic pitch names.
For example, the following command will transpose down an augmented unison
(e.g. C\# to C):

\par 

\texttt{trans -d 0 -c -1 input > output}



\par 
Conversely, we can respell the diatonic pitches without changing the overall
pitch height.
For example, the following transposition will transpose "up"
a diminished second (e.g. from F-sharp to G-flat):

\par 

\texttt{trans -d +1 -c 0 input > output}



\par 
Modal transpositions are also possible by omitting the chromatic offset
option.
Consider, for example, the following C major scale:

\texttt{**kern
d
e
f
g
a
b
cc
*-}

We can transform this using the following diatonic transposition:

\par 

\texttt{trans -d +1}


\par 
The resulting output is the Dorian mode:

\texttt{**kern
*Trd1
d
e
f
g
a
b
cc
dd
*-}

When using the
\textbf{-d}
option alone,
\textbf{trans}
eliminates all accidentals in the input.
This can be potentially confusing, but it is often useful.
Suppose you have a passage in the key of E major which
you would like to translate to E Dorian.
First transpose so the tonic is D using only the
\textbf{-d}
option; then transpose exactly so the tonic is E again:

\par 

\texttt{trans -d -1 Emajor | trans -d +1 -c +2 > Edorian}


\par 
For some changes of mode (such as melodic to harmonic minor),
you may need to use the
\textbf{humsed}
command described in
Chapter 14
to modify accidentals for specific
scale degrees.

\par 
Notice the addition of a "tandem interpretation" to the
above example (\texttt{*Trd1}).
Whenever
\textbf{trans}
is invoked, it adds a record indicating
that the encoding is no longer at the original pitch.
\textit{Transposition tandem interpretations}
are similar in syntax to the
\textbf{trans}
command itself.
In the above example, \texttt{*Trd1} indicates a diatonic shift up
one letter name.
The tandem interpretation \texttt{*Trd-1c-2}
would indicate that a score has been transposed down a major second.
The
\textbf{trans}
command also provides a
\textbf{-k}
option that allows the user to specify a replacement key signature
for the output.

\par 
The
\textbf{trans}
command can be used in conjunction with any of the appropriate
pitch-related representations, such as \texttt{**pitch}, \texttt{**kern},
\texttt{**Tonh}, and \texttt{**solfg}.


\subsection*{Key Interpretations}

\par 
In order for the
\textbf{solfa}
or
\textbf{deg}
commands to translate from other pitch representations,
the encoded music must contain an explicit key indication.
Keys are explicitly represented by a single asterisk, followed by
an upper- or lower-case letter, followed by an optional
accidental, followed by a colon.
The octothorpe (\texttt{\#}) indicates a sharp and the
hyphen (\texttt{-}) indicates a flat.

\par 
Upper-case letters indicate major keys;
lower-case letters indicate minor keys.
By way of illustration, the following key interpretations
indicate the keys of C major, C minor, B-flat major,
and F-sharp minor:

\texttt{*C:
*c:
*B-:
*f\#:


\par 
Key interpretations usually appear near the beginning of
a representation, and key interpretations can be redefined
at any place in a score.


\subsection*{Pitch Processing}

\par 
Apart from transposition, 
translating from one representation to another provides
opportunities for different sorts of processing.
Suppose, for example, we wanted to know whether the subdominant
pitch occurs more frequently in one vocal repertory than in
another repertory.
We can use
\textbf{solfa}
in conjunction with \textbf{grep}'s
\textbf{-c}
option to count the number of occurrences.
(For the following examples, we will assume that the inputs
consist of only a single spine, that barlines are absent,
and that appropriate interpretations are provided indicating
the key of each work.)
First we need to count the total number of notes in each repertory.

\par 

\texttt{census -k repertory1.krn}
\\
\texttt{census -k repertory2.krn}


\par 
Next we translate the scores to the
\textbf{solfa}
representation and use
\textbf{grep -c}
to count the number of occurrences of the number `fa':

\par 

\texttt{solfa repertory1.krn | grep -c fa}
\\
\texttt{solfa repertory2.krn | grep -c fa}


\par 
The proportion of subdominant pitches can be calculated by
simply comparing the resulting pattern count with the number
of notes identified by
\textbf{census}.

\par 
Recall that one of the differences between the
\texttt{**solfa}
and
\texttt{**deg}
representations is that the \texttt{**deg}
output contains an indication of the direction of melodic
approach.
The caret (\textbf{\^{}}) indicates approach from below, whereas
the lower-case \textbf{v} indicates approach from above.
Suppose we wanted to determine whether the dominant pitch
is more commonly approached from above or from below.
Assuming a monophonic input, we can once again use
\textbf{grep}
to answer this question.
First let's count how many dominant pitches (`5') are approached
from above (`v'):

\par 

\texttt{deg repertory.krn | grep -c v5}


\par 
The caret has a special meaning for
\textbf{grep}
which will be discussed in
Chapter 9.
We can escape the special meaning by preceding the caret by
a backslash.
In order to count the number of dominant pitches approached
from below we can use the following:

\par 

\texttt{deg repertory.krn | grep -c $\backslash$\^{}5}


\par 
Recall that some scale tones are spelled differently depending 
on whether the mode is major or minor.
For example, in A major the mediant pitch is C sharp; but in A minor
the mediant pitch is C natural.
The
\textbf{deg}
and
\textbf{solfa}
commands produce subtly contrasting outputs that make one or
the other command better suited depending on the user's goal.
The
\textbf{deg}
command would represent C sharp in A major, and C natural in A minor
by the same scale degree -$\,$- \texttt{3}.
In the key of A major, C natural would be characterized as a
lowered mediant (\texttt{3-}) and in A minor, C sharp would be
characterized as a raised mediant (\texttt{3+}).
By contrast, the
\textbf{solfa}
command characterizes pitches with respect to the tonic alone
and ignores the mode.
Hence,
\textbf{solfa}
would designate C sharp as `\texttt{mi}' whether the key was A major or
A minor.
Similarly, C natural would be designated `\texttt{me}' in both A major
and A minor.
The differences between
\textbf{deg}
and
\textbf{solfa}
allow users to distinguish chromatically altered scale tones in
a manner appropriate to the task.


\subsection*{Uses for Pitch Translations}

\par 
Occasionally it is useful to process a given representation
to the
\textit{same}
representation.
The
\textbf{kern}
command translates various pitch-related representations to the \texttt{**kern}
format.
The
\textbf{-x}
option eliminates any input data that do not pertain to pitch.
When applied to a \texttt{**kern} input, this option allows us to
filter out durations, articulation marks,
phrasing, and other non-pitch data.
Suppose, for example, that we wanted to determine the proportion
of successively repeated notes in a vocal melody:
how often is a pitch followed immediately by the same pitch?
We might begin by first determining the total number of notes in the melody
using \textbf{census} with the \textbf{-k} option.

\par 

\texttt{ census -k melody.krn}


\par 
We can use the
\textbf{uniq}
command to eliminate successive repeated pitches -$\,$- but only
if the note tokens are identical.
First we can use
\textbf{kern -x}
to translate "from \texttt{**kern} to \texttt{**kern}" while
eliminating non-pitch-related data.
Then we need to remove barlines so they don't interfere with
pitches that are repeated across the measure.
Using
\textbf{uniq}
will then eliminate all of the successively duplicated records,
so a sequence of six G's will be reduced to a single G.
Finally, we pipe the output to
\textbf{census -k}
to count the total number of notes.

\par 

\texttt{kern -x melody.krn | uniq | census -k}



A variation on this approach would entail translating to a
representation that does not distinguish enharmonic pitches.
For example, translating our melody to
\texttt{**semits}
and
then back to
\texttt{**kern}
will standarize all of the enharmonic spellings.
If our melody contains a G-sharp that undergoes an enharmonic
shift to A-flat, then the pitches will be deemed identical.
The following command carries out the same task as above,
but ignores possible enharmonic spellings:

\par 

\texttt{semits melody.krn | kern | uniq | census -k}


\par 
Incidentally, given \texttt{**semits} input, the
\textbf{kern}
command will spell pitches according to any key or key signatures
it encounters.
For example, if the key signature contains sharps, then G-sharp
will be output;
if the key or key signature contains flats,
then A-flat will be output.



\subsection*{Reprise}

\par 
In this chapter we have introduced a number of pre-defined
pitch-related representations.
Simple commands can be used to translate from one representation to another.
Which representation is most appropriate depends on the user's goal.

\par 
There is a wealth of other representation formats related to
pitch distances, tablatures, timing, and other types of musical information.
These representations will be explored in later chapters.
In addition, we'll describe how to design your own representations
-$\,$-  representations that may be better tailored to a specific application.
However, before we continue discussing further representations,
this is an appropriate point to present a more formal description
of the general Humdrum representation syntax.

\\
\begin{itemize}
\item 

\textbf{Next Chapter}
\item 

\textbf{Previous Chapter}
\item 

\textbf{Table of Contents}
\item 

\textbf{Detailed Contents}
\\\\

© Copyright 1999 David Huron
\end{document}
